Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-04-05 Chapter-1/1-Theory.tex
s

The first two lines of the lagrangian summarize the kinetic terms for the fermionic fields and their coupling to the gauge bosons $A^{\mu\nu}_a$, $B^{\mu\nu}$ \footnote{In gauge theories the ordinary derivative $\partial_\mu$  is substitued with ¬èthe covariant derivative $D_\mu$. Here $D_\mu = \partial_\mu + igA_\mu \cdot I + ig'B_\mu\frac{Y}{2}$, where I and Y are the SU(2)$_L$ and U(1)$_Y$ generators, respectively.}.
The third line describes the kinetic terms and the self-coupling terms of the gauge bosons. The forth line is the Higgs lagrangian, which results in the spontaneous symmetry breaking. The last three lines describe the Yukawa coupling between fermions and the Higgs field, origin of the fermion's mass.

@@

where $c$ e $s$ stand respectively for cosine and sine of the corresponding  mixing angles ($\theta_{12}$, $\theta_{23}$ and $\theta_{13}$), $\delta$ èis the Dirac CP violation phase, $\alpha_1$ and $\alpha_{2}$   is the eventual Majorana CP violation phases.  Experimental results on neutrino oscillations are generally reported in terms of the mixing angles and of the squared mass splitting $\Delta m^2_{ab} = | m^2_{a} - m^2_{b}|$, where $a$ and $b$ represent the mass eigenstates. A summary of the current status of experimental results, albeit partial, is given in table \ref{tab:nuosc}.

@@

\subsubsection{Open Questions in Neutrino Physics}
On one hand, the last \textcolor{red}{X} decades of experiments in neutrino oscillations brought spectacular advancements in the understanding of the oscillations pattern, measuring the neutrino mixing angles and mass splitting with a precision of less than 10\%.  On the other, it opened the field for a series of questions needing experimental answers. 
Critical challenges in the next decade will entail an even deeper understanding of the neutrino mixing pattern, investigation of the neutrino mass origin, neutrino number and nature, and assessing CP violation in the lepton sector.

@@

\textcolor{red}{Smallness of neutrino masses, link to GUTs}

@@

@@

>>>>

The first two lines of the lagrangian summarize the kinetic terms for the fermionic fields and their coupling to the gauge bosons $A^{\mu\nu}_a$, $B^{\mu\nu}$ \footnote{In gauge theories the ordinary derivative $\partial_\mu$  is substitued with Âthe covariant derivative $D_\mu$. Here $D_\mu = \partial_\mu + igA_\mu \cdot I + ig'B_\mu\frac{Y}{2}$, where I and Y are the SU(2)$_L$ and U(1)$_Y$ generators, respectively.}.
The third line describes the kinetic terms and the self-coupling terms of the gauge bosons. The forth line is the Higgs lagrangian, which results in the spontaneous symmetry breaking. The last three lines describe the Yukawa coupling between fermions and the Higgs field, origin of the fermion's mass.

@@

where $c$ e $s$ stand respectively for cosine and sine of the corresponding  mixing angles ($\theta_{12}$, $\theta_{23}$ and $\theta_{13}$), $\delta$ is the Dirac CP violation phase, $\alpha_1$ and $\alpha_{2}$   is the eventual Majorana CP violation phases.  Experimental results on neutrino oscillations are generally reported in terms of the mixing angles and of the squared mass splitting $\Delta m^2_{ab} = | m^2_{a} - m^2_{b}|$, where $a$ and $b$ represent the mass eigenstates. A summary of the current status of experimental results, albeit partial, is given in table \ref{tab:nuosc}.

@@

\subsubsection{Open Questions in Neutrino Physics}
On one hand, the last \textcolor{red}{X} decades of experiments in neutrino oscillations brought spectacular advancements in the understanding of the oscillations pattern, measuring the neutrino mixing angles and mass splitting with a precision of less than 10\%.  On the other, it opened the field for a series of questions needing experimental answers. 
Critical challenges in the next decade will entail an even deeper understanding of the neutrino mixing pattern, investigation of the neutrino mass origin, the determination of the number of neutrinos and their nature, and the measurement of CP violation in the lepton sector.

@@

The smallness of neutrino masses remains a theoretical puzzle.
\textcolor{red}{Smallness of neutrino masses, link to GUTs}

@@

\subsubsection{Nucleon decay}\label{sec:theoryPDK}
Baryon number is accidentally conserved in the Standard Model. Even though no baryon number violation has been experimentally observed thus far, no underlying symmetry in line with the Noether paradigm \cite{Noether1971} explains its conservation. Almost all Grand Unified Theories predict at some level baryon number violation in the form of nucleon decay on long time-scales.  Given the impossibility to reach grand unification energy scales with collider experiments ($\sqrt{s} > 10^{15}$ GeV),  an indirect proof of GUT is needed. The experimental observation of nucleon decay may be the only viable way to explore these theories and it is therefore a subject of great interest \cite{Adams:LBNE}. %Both experiments and theory indeed suggest the energy scale for convergence of the running coupling constants of the Standard Model to be over $10^{15}$ GeV. This energy scale seems impossible to access by any foreseeable accelerator experiment, leaving baryon number violation  to be the only testable process.

@@

\subsection{Pion-Argon Total Hadronic Cross Section}
\textcolor{red}{Here we put the link to previous sections: why is this important in the current context?}
\subsubsection{Previous measurements}
\subsubsection{Signal Signatures}
\subsection{Kaon-Argon Total Hadronic Cross Section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%% ---------------------------------------- Hadronic Cross sections ---------------------------------------- %%%%%

\subsubsection{K$^{+}$Ar Cross section in the Context of Nucleon Decay Searches}
If nucleon decay was experimentally found, additional information about the GUT type may be derived from the dominant decay mode.  Supersymmetric GUTs \cite{Dimopoulos:1981dw,Bajc20161} prefer the presence of kaons in the products of the decay, e.g. $p\rightarrow K^+\bar{\nu}$  (see fig \ref{fig:MandatoryFeynmannDiagrams}, left).
Gauge mediation GUTs, in which new gauge bosons are introduced that allow for the transformation of quarks into leptons, and vice versa, prefer the mode $p\rightarrow e^+\pi^0$ (see fig \ref{fig:MandatoryFeynmannDiagrams}, right).
%%%%%%%%%%%%%%%% Find theory papers!!!!!! %%%%%%%%%%

\begin{figure}[hbpt]
\centering
\includegraphics[width=6.5in]{Chapter-1/Images/MandatoryFeynmannDiagrams.png}
\caption{Feynman diagrams for proton decay ``golden modes": $p \rightarrow K^+ \bar{\nu}$ for supersymmetric GUTs on the left and  $p \rightarrow e^+ \pi^0$ for gauge-mediation GUTs  on the right.}
\label{fig:MandatoryFeynmannDiagrams}
\end{figure}

LArIAT tiny active volume makes it impossible for the experiment to place competitive limits on nucleon decay searches.  However,  LArIAT provides excellent data to characterize kaons in LAr for the ``LAr golden mode", $p \rightarrow K^+ \bar{\nu}$. The result of these studies will affect future proton decay searches in LArTPCs. Previous work has been done to assess the potential identification efficiency for different decay modes in a LArTPC \cite{Bueno2007}, but, as the time of this  writing, no study of kaon selection efficiency in LArTPCs has been performed on data. 
The K$^+$-Ar interaction cross section has never been measured before and can affect the possibility of detecting and measuring kaons when produced in a proton decay event. 
Kaon interactions with argon can distort the kaon energy spectrum as well as change the topology of single kaon events. In a LArTPC, non-interacting kaons appear as straight tracks with a high ionization depositions at the end (Bragg peak). The topology of interacting kaons can be quite different. In case of elastic scattering, a distinct kink will be present in the track. In case of inelastic scattering the Bragg peak will not be present and additional tracks will populate the event.
Performing the total hadronic K$^+$-Ar cross section measurement on data serves the double purpose of identifying the rate of ``unusual" topologies (kinks and additional tracks) and of developing tools for kaon tracking in LAr.

\subsubsection{K$^{+}$Ar Cross section in the Context of Light Mesons Interaction with Nuclei}
\label{sec:theoryStrangeMeson}
The intrinsic value of this measurement is that kaon interactions complement the measurements of $\pi$ interactions as a probe of  hadron interaction inside the nucleus in the strange sector.  
\textcolor{red}{CHIEDI REFERENZE A FLAVIO}

\subsubsection{Signal Signatures}\label{sec:KSignalSignature}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The interaction of a mildly relativistic charged kaon with an argon nucleus is determined largely by the strong force. The total hadronic K$^{+}$-Ar interaction cross section is defined as the one related to the single (hadronic) process driven only by the strong interaction.
In this case, ``total" indicates all strong interactions regardless of the final state. This condition purposefully includes both elastic and inelastic (reaction) channels. Indeed, the total cross section section can be then decomposed into
$$\sigma_{Tot} = \sigma_{Elastic}+ \sigma_{Reaction}.$$

For this analysis, kaons are selected from the LArIAT beamline in the momentum range between \textcolor{red}{500} MeV/c and \textcolor{red}{1000} MeV/c (see Fig \ref{fig:TOFK}).

\begin{figure}[hpbt]
\centering
%\includegraphics[width=5in]{Chapter-1/Images/KaonTOF}
\caption{Time of flight versus momentum distributions as produced by the LArIAT TOF and Wire Chambers systems. The Kaon population lies between the proton and the muon/pion populations, allowing PID of Kaons in the beam line.  }
\label{fig:TOFK}
\end{figure}

In this energy range, the relevant K-Nucleon interactions are according to \cite{fesbach1992theoretical}:
\begin{equation}
K^{+} + N \rightarrow K^{+} + N\textit{ (elastic)}
\end{equation}
\begin{equation}K^{+} + n \rightarrow K^{0} + p\textit{ (elastic)}\end{equation}
\begin{equation}K^{+} + N \rightarrow K + N + \pi \textit{ (inelastic)}\end{equation}
\begin{equation}K^{+} + N \rightarrow K^{*} + N\textit{ (inelastic)}.\end{equation}

\subsubsection{Previous Measurements: Lighter and Heavier Nuclei}
In general, data on kaon cross sections are  extremely scarce. The measurement of kaon interaction cross section would bring the additional benefit of reducing the uncertainties associated  with hadron interaction models adopted in MC simulations for argon targets.

The data on nuclear effects for specific hadronic final states are extremely scarce, resulting in big uncertainties in modeling final state interactions \cite{Drakoulakos:2004gn}. Figure \ref{fig:Friedmann} shows a 1997 measurement on several elements as performed by  Friedmann et al.  \cite{Friedman:1997eq}. As a reference, this paper measures a $\sigma_{Tot}$ for Si of  366.5  $\pm$  4.8 mb and a $\sigma_{Tot}$ for Ca of 494.6  $\pm$ 7.7 mb at 488 MeV/c.  The cross section for argon is expected to lie in between these two measurements. 
Additional data on the kaon cross section are provided by Bugg et al. \cite{PhysRev.168.1466}. Bugg performs a measurement of the total 
K$^+$ and K$^-$ cross sections on protons and deuterons over the range of 0.6-2.65 GeV/c, as well as a measurement of the total K$^+$ and K$^-$  cross sections on carbon for a number of momenta; the results of this paper on carbon are reported in Figure \ref{fig:Bugg}.

\begin{figure}
\begin{minipage}[b]{.5\textwidth}  
  \centering  
\includegraphics[width=3in]{Chapter-1/Images/Friedmann.png}
\end{minipage}%  
\begin{minipage}[b]{0.5\textwidth}  
  \centering  
\includegraphics[width=3in]{Chapter-1/Images/Bugg.png}
\end{minipage}
\par
\begin{minipage}[t]{.53\textwidth}
\caption{Ratios between experimental and calculated cross sections as from \cite{Friedman:1997eq}. Top: Total cross sections. Bottom: reaction cross sections.}
\label{fig:Friedmann}
\end{minipage}%
\begin{minipage}[t]{.5\textwidth}  
\caption{Total K$^+$  and K$^-$ cross sections on carbon as from \cite{PhysRev.168.1466}.}
\label{fig:Bugg}
\end{minipage}  
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% PRETTY EVENT DISPLAY WITH TEXT, NOT SURE IF USEFUL
%\begin{figure}[h!]
%\centering
%\includegraphics[width=6.5in]{Chapter-1/Images/KLariat.png}
%\caption{LArIAT Data $K^+$ candidate. $K^+$ enters TPC, undergoes a hadronic scatter, and then decays into $\pi^+$ and $\pi^0$. The the $\pi^0$ decays into 2 photons while the $\pi^+$ stops quickly in the TPC. Collection plane view.}
%\label{Fig:KLariat}
%\end{figure}

%Fig \ref{Fig:KLariat} shows a $K^+$ candidate event in the LArIAT TPC. Following the kaon candidate track from left to right, two important elements are visible by eye: a change in the K momentum due to hadronic scatter and a Bragg peak by the end of the track due to an augment of ionization as the kaon slows down in the TPC. The track "kink" is only visible thanks to the millimetric spacial resolution of the TPC, while the Bragg shows the calorimetric power of this technology. The kaon in this event decays hadronically into $\pi^+$ and $\pi^0$. The the $\pi^0$ decays into 2 photons while the $\pi^+$ stops quickly in the TPC. The ability to distinguish the topology of this decay from the most frequent one, i.e. $K^+\rightarrow\mu^+\nu$, remarks the versatility of the LArTPC technology.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 \subsubsection{Kaon Interaction Cross Section for thin target in Geant4}
Since the kaon cross section in argon has never been measured before, the Geant4 Monte Carlo tunes kaon transportation in argon by extrapolation from lighter and heavier nuclei. As shown in the previous section,  kaon data on carbon are available and  can be used as a metric to evaluate the Geant4 prediction performances.  Figure \ref{fig:TrueCarbon} shows the total hadronic cross section for carbon implemented in Geant4 10.01.p3 overlaid with the Bugg and Friedman data. Unfortunately, the current version of Geant4 does not reproduce the data for carbon closely. On one hand, this evidence makes us even more wary when using the Monte Carlo in simulating the kaon-argon interactions. On the other, it further highlights the importance of kaon measurements.

### 2018-04-05 Chapter-2/2-LArTPCs.tex
s

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\section{Enabling the next generation of discoveries: LArIAT}
LArIAT, a small Liquid Argon Time Projection Chamber (LArTPC) in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. LArTPC represents one of the most advanced experimental technologies for physics at the Intensity Frontier due to its full 3D-imaging, excellent particle identification and precise calorimetric energy reconstruction. This complex technology however needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries at the Intensity Frontier which LArIAT can provide.

>>>>

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold.  Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying exclusively on photons from nuclear de-excitation. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a Liquid Argon Time Projection Chamber (LArTPC).

LArTPCs can complement nucleon decay searches in modes where water Cherenkov detectors are less sensitive, especially $p\rightarrow K^+\bar{\nu}$. According to \cite{Acciarri:Dune}, DUNE will have an active volume large enough, have sufficient shielding from the surface, and will run for lengths of time sufficient to compete with Hyper-K, opening up the opportunity for the discovery of nucleon decay.

\begin{figure}[hbpt]
\centering
\includegraphics[width=6.5in]{Chapter-2/Images/PDKExperimentalLImit.png}
\caption{Proton decay lifetime limits from passed and future experiments.}
\label{fig:PDKExperimentalLImit}
\end{figure}

%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\section{Enabling the next generation of discoveries: LArIAT}
LArIAT, a small Liquid Argon Time Projection Chamber (LArTPC) in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. LArTPC represents one of the most advanced experimental technologies for physics at the Intensity Frontier due to its full 3D-imaging, excellent particle identification and precise calorimetric energy reconstruction. This complex technology however needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries at the Intensity Frontier which LArIAT can provide.
